---
title: "Week 4 Activity, Part 1"
author: 'Rithu Uppalapati'
date: "9/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part 1 (dplyr)

In the part 1 of the activity you will write or interpret short dplyr pipes
that explore or analyze the `Orange` data from R's `datasets` package. The data
which contains 35 rows and 3 columns recording the growth
of orange trees.  The three columns in the data set are:

  - `Tree`: an ordered factor, identifying individual trees, 
  - `age`: a numeric vector giving the number of days 
    since the tree was planted,
  - `circumference`: a numeric vector recording the circumference of the 
    trunk in mm. 

You can read more at `?Orange`. 

1. Write a dplyr pipe to determine the number of
observations per tree.

```{r q1}
Orange %>%
  group_by(Tree) %>%
  summarize(n_obs = n(), .groups = 'drop') %>% 
  group_by(n_obs) %>%
  summarize(n_trees = n(), .groups = 'drop')
```

2. Write a dplyr pipe to change the units of age to "years" 
and circumference to "cm".

Note the change in capitalization for version in global environment. 

```{r q2}
orange = datasets::Orange %>%
  mutate( 
   age = age / 365, # could use 365.25 as well
   circumference = circumference / 10
  )
```

 
3. Write a dplyr pipe to add a column assigning a
   z-score to each tree, centered around the mean *circumference* 
   for all trees at a given age.

```{r q3}
orange = orange %>%
  group_by(age) %>%
  mutate(z_circ = {circumference - mean(circumference)} / sd(circumference))
```

4. Write a dplyr pipe to calculate the average rate of growth
(cm/year) across all trees between between age 0 (assume circumference = 0) and
the final measured age for each tree.


```{r q4}
orange %>%
  group_by(Tree) %>%
  filter( age == max(age) ) %>%
  summarize( growth_rate = mean(circumference / age) )
```

5. Describe the result of the following pipe in 1-3 sentences. Your
    *description* should touch on both the rows and columns and also describe
    a substantive question addressed by the result. 


```r
 orange %>%
   group_by(Tree) %>%
   mutate( 
     new_growth = c(circumference[1], diff(circumference)),
     elapsed = c(age[1], diff(age))
   ) %>%
   group_by(age) %>% 
   summarize( 
     avg_rate = mean(new_growth / elapsed),
     se = sd( new_growth / elapsed ) / sqrt(n()) 
   )
```

**Description:** The result of this pipe is the average growth of all tried over the elapsed age. 
Our starting point is 0 cm circumfrence and 0 year old tree. The first row correspond to the average rate
of groimplicitly assuming that the circumference was 0 cm at age 0 years. 
Each row corresponds to average growth rate and the second row corresponds to the average error. 
